---
title: "Creating a map with FishMIP regional models"
author: "Denisse Fierro Arcos"
date: "2023-11-01"
output: 
  github_document:
    toc: true
    html_preview: false
---

# Introduction

Now that we have a single shapefile with all the FishMIP regional models, we will create a map that can be used in any publication.  
  
Here, we will show how to access and download the FishMIP regional models shapefile available from the THREDDS server.  
  
# Loading libraries
  
```{r libraries}
#Connection to THREDDS
library(thredds)
#Downloading data
library(curl)
#Manipulating data
library(dplyr)
#Spatial data
library(sf)
#Base map
library(rnaturalearth)
#Plotting maps
library(ggplot2)
library(cowplot)
library(RColorBrewer)
```

# Downloading FishMIP regional models shapefile from THREDDS
We will use the `HTTPServer` service to download the shapefile files which are available as a `zip` folder.  
  
**Note** that you only need to download shapefiles once.  
  
```{r download_thredds, eval = FALSE}
#Defining domain name for THREDDS server
domain <- "http://portal.sf.utas.edu.au"
#Defining location of regional shapefile in THREDDS server
fishmip_reg_cat <- paste0(domain, "/thredds/catalog/gem/fishmip/FishMIP_regions/catalog.xml")

#Getting catalog items available in server
fishmip_reg <- CatalogNode$new(fishmip_reg_cat, prefix = "thredds")

#Getting base url for HTTPServer downloads
base_http <- paste0(domain, fishmip_reg$list_services()$http[["base"]])

#Getting names of datasets available for download
ds_name <- fishmip_reg$get_dataset_names()

#Getting complete download URL
url <- paste0(base_http, fishmip_reg$get_datasets()[[ds_name]]$url)

#Downloading zip file
curl_download(url, destfile = "FishMIP_regions.zip")

#Unzipping folder - storing files in Output folder
out_folder <- "../Outputs"
#If folder does not exist, create one
if(!dir.exists(out_folder)){
  dir.create(out_folder, recursive = T)}
unzip("FishMIP_regions.zip", exdir = out_folder)

#Delete zip folder
unlink("../FishMIP_regions.zip")

#Removing all variables from environment
rm(list = ls())
```

# Loading FishMIP regional models shapefile
Now that we have downloaded the merged shapefiles, we can load it to `R` and create a map. We will exclude the Southern Ocean region.  
  
```{r}
fishmip_reg <- read_sf("../Outputs/FishMIP_regional_models/FishMIP_regional_models.shp") |> 
  filter(region != "Southern Ocean")

#Checking results
fishmip_reg
```
  
# Plotting map
Since we removed the Southern Ocean region, we have 22 polygons left. However, we will need to apply some changes to our datasets to create publication ready maps.  
  
## Data preparation  
We will use a Robinson projection for our map, so we need to reproject our shapefile. A reprojection also needs to be applied to our base world map.  
  
We will also create a an original colour map to plot the FishMIP regions using a combination of `RColorBrewer` palettes. We will add this information to the reprojected regional shapefile for plotting.
  
```{r}
#Define Robinson projection to be applied to shapefile and basemap
rob_proj <- "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m +no_defs +type=crs"

#Load world base map
world <- ne_countries(scale = "medium", returnclass = "sf") |> 
  #Ensuring any polygons crossing international dateline are correctly plotted
  st_wrap_dateline() |> 
  #Reprojecting
  st_transform(rob_proj)

#Creating a colour palette by merging colour brewer palettes
pal <- c(brewer.pal(8, "Set2"), brewer.pal(12, "Paired"))

#Reprojecting FishMIP regions
fishmip_reg_rob <- fishmip_reg |> 
  #Adding colour palette
  mutate(fill = pal) |> 
  st_wrap_dateline() |> 
  st_transform(rob_proj)
```
  
## Mapping FishMIP regional models
Since there is some overlap in the spatial coverage for regional models, we will create maps showing more detail in the overlapping regions.  
  
First, we will plot all regions and extract the legend.   
  
```{r}
#Plotting 
reg <- fishmip_reg_rob |> 
  #Plotting
  ggplot()+
  #Using colours of our tailored made colourmap
  geom_sf(aes(fill = region), alpha = 0.5)+
  geom_sf(data = world, fill = "#f9f9f5", show.legend = F)+
  scale_fill_manual(values = fishmip_reg_rob$fill)+
  theme_bw()+
  #Move legend to bottom
  theme(legend.position = "bottom", legend.title = element_blank(),
        legend.margin = margin(0, 0, 0, 0),
        legend.box.margin = margin(0, 0, 0, 0),
        legend.text = element_text(margin = margin(r = 5)),
        panel.border = element_rect(colour = NA))+
  #Split legend into four columns
  guides(fill = guide_legend(ncol = 4))

#Checking map so far
reg
```
  
Our initial maps shows all FishMIP regions, but does not look great yet. We will use this as a guide to identify the areas where we will create the smaller maps. We will expand the following areas:  
- Europe  
- Brazil  
- Australia and New Zealand  
- East Antarctica  
  
We will use the base map above to create these smaller maps.  

### Europe
  
```{r}
europe <- reg+
  #We will increase transparency to see overlapping areas better
  geom_sf(inherit.aes = T, aes(alpha = 0.3))+
  #Remove background
  theme_bw()+
  #Remove legend
  theme(legend.position = "none")+
  #Add world base map
  geom_sf(data = world, fill = "#f9f9f5")+
  #Focus on Europe. Note that we use coordinates in meters because data is reprojected
  lims(x = c(-741458, 3251458), y = c(3256289, 6725154))+
  #Add a border to map so it is easily identifiable
  theme(panel.border = element_rect(colour = "#0077bb", linewidth = 2),
        axis.text = element_blank(), axis.ticks = element_blank(),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"))

#Create a shapefile with map limits
eu_box <- st_bbox(c(xmin = -751458, xmax = 3351458, ymax = 6825154, ymin = 3156289), 
                  crs = rob_proj) |> 
  st_as_sfc()

#Check map of Europe
europe
```
  
### Brazil
  
```{r}
brazil <- reg+
  #We will increase transparency to see overlapping areas better
  geom_sf(inherit.aes = T, aes(alpha = 0.3))+
  #Remove background
  theme_bw()+
  #Remove legend
  theme(legend.position = "none")+
  #Add world base map
  geom_sf(data = world, fill = "#f9f9f5")+
  #Focus on Brazil.
  lims(x = c(-3537222, -3151458), y = c(-1206289, -400000))+
  #Add a border to map so it is easily identifiable
  theme(panel.border = element_rect(colour = "#cc6677", linewidth = 2),
        axis.text = element_blank(), axis.ticks = element_blank())

#Create a shapefile with map limits
br_box <- st_bbox(c(xmin = -3537222, xmax = -3151458, ymax = -400000, ymin = -1206289),
                  crs = rob_proj) |>
  st_as_sfc()

#Check map of Brazil
brazil
```
  
  
### Australia and New Zealand
  
```{r}
au_nz <- reg+
  #We will increase transparency to see overlapping areas better
  geom_sf(inherit.aes = T, aes(alpha = 0.3))+
  #Remove background
  theme_bw()+
  #Remove legend
  theme(legend.position = "none")+
  #Add world base map
  geom_sf(data = world, fill = "#f9f9f5")+
  #Focus on Australia and New Zealand
  lims(x = c(10568611, 15500000), y = c(-5112577, -2600000))+
  #Add a border to map so it is easily identifiable
  theme(panel.border = element_rect(colour = "#999933", linewidth = 2),
        axis.text = element_blank(), axis.ticks = element_blank(),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"))

#Create a shapefile with map limits
au_nz_box <- st_bbox(c(xmin = 10468611, xmax = 15600000, ymax = -2500000, ymin = -5212577),
                  crs = rob_proj) |>
  st_as_sfc()

#Check map
au_nz
```
  
### Southern Ocean
  
```{r}
# so <- 
reg+
  #We will increase transparency to see overlapping areas better
  geom_sf(inherit.aes = T, aes(alpha = 0.3))+
  #Remove background
  theme_bw()+
  #Remove legend
  theme(legend.position = "none")+
  #Add world base map
  geom_sf(data = world, fill = "#f9f9f5")+
  #Focus on Australia and New Zealand
  lims(x = c(4198611, 6800000), y = c(-7225154, -4712577))+
  #Add a border to map so it is easily identifiable
  theme(panel.border = element_rect(colour = "#332288", linewidth = 2),
        axis.text = element_blank(), axis.ticks = element_blank(),
        plot.margin = margin(0, 0, 0, 0, unit = "cm"))

#Create a shapefile with map limits
so_box <- st_bbox(c(xmin = 4098611, xmax = 6900000, ymax = -4612577, ymin = -7325154),
                  crs = rob_proj) |>
  st_as_sfc()

#Check map
so
```
  
### Adding boundaries of inset maps into main map
  
```{r}
main <- reg+
  geom_sf(data = eu_box, color = "#0077bb", fill = NA, linewidth = 0.5)+
  geom_sf(data = au_nz_box, color = "#999933", fill = NA, linewidth = 0.5)+
  geom_sf(data = so_box, color = "#332288", fill = NA, linewidth = 0.5)

main
```



```{r}
# inset_left <- 
plot_grid(NULL, main, plot_grid(NULL, europe, au_nz, so, NULL, NULL, nrow = 6), 
          ncol = 3, rel_widths = c(0.1, 1., 0.25))

ggsave("../test.pdf", device = "pdf")

```
### Merging main map with inset maps
  
```{r}
#Merging main map and insets
main_inset <- ggdraw(main)+
  draw_plot(au_nz, x = .75, y = 0.125, width = 0.25, height = 0.3)+
  draw_plot(so, x = .75, y = 0.45, width = 0.225, height = 0.25)+
  draw_plot(europe, x = .74, y = 0.75, width = 0.25, height = 0.25)

ggsave("../test.pdf", main_inset, device = "pdf", width = 15, height = 7, units = "cm")

#Adding legend
final_map <- main_inset+
  theme(plot.margin = margin(b = 3.25, t = -2.5, unit = "cm"))+
  draw_plot(leg, x = 0, y = -0.475)

#Checking final result
final_map
```
  
The map above may not look great in your screen, but once it is saved to your local machine, it will plot correctly. However, you can change the code here to fit it to your needs.    
  
## Saving final map
The final map can be saved to your local machine. We will save it as vector image so it still looks good even when expanded. You can also find a copy of this image in the [FishMIP THREDDS server](http://portal.sf.utas.edu.au/thredds/catalog/gem/fishmip/catalog.html).  
  
```{r, eval = F}
#Saving map
ggsave("../Outputs/FishMIP_regional_models.svg", final_map, device = "svg")
```


