---
title: "Creating_raster_mask"
author: "Denisse Fierro Arcos"
date: "2023-11-06"
output:
  github_document:
    toc: true
    html_preview: false
---

## Introduction

In this notebook, we will show how to create a raster mask that can be used to extract data from Earth System Models. Here, we will use the shapefile for the Southern Ocean MICE regional model. This regional model contains some internal boundaries that do not overlap. We will create two-dimensional masks that will match the various grids used in the Fish-MIP project.  
  
# Loading libraries
  
```{r libraries}
knitr::opts_chunk$set(fig.path = "figures/")
#Spatial data
library(sf)
library(terra)

#Manipulating and plotting data
library(tidyverse)

#Base map
library(rnaturalearth)
```
  
# Loading Southern Ocean shapefile
Given that some subregions cross the international date line, the shapefile contains two polygons corresponding to the same region. We will assign a unique ID for each region, which will ensure that if there are two polygons for a region, they will be classified as that region.  
  
```{r}
#Loading shapefile
southern_ocean <- file.path("/rd/gem/private/shared_resources/",
                            "Shapefiles_Regions/Southern-Ocean_MICE",
                            "SupportInfo/SouthernOcean_MICE.shp") |> 
                            read_sf() |> 
  #Uniting region and band columns
  unite("region", region, band)

#Create a unique ID for each region
so_id <- southern_ocean |> 
  distinct(region) |> 
  rowid_to_column("id")

#Adding unique ID to shapefile
southern_ocean <- southern_ocean |> 
  left_join(so_id, by = "region")

#Checking result
ne_countries(returnclass = "sf") |> 
  ggplot()+
  geom_sf()+
  geom_sf(inherit.aes = F, data = southern_ocean, aes(fill = region), 
          alpha = 0.7)+
  theme_bw()
```
  
The Southern Ocean regions are plotting correctly, now we can move onto creating raster masks. In total, we will create four different masks to match all inputs and outputs used in Fish-MIP.  
  
# Finding location of raster samples
We have a folder containing samples of the raster used in Fish-MIP models and ESMs. We will list all the files contained in that folder.    
  
```{r}
#Directory containing sample rasters
samples_dir <- "/rd/gem/private/shared_resources/grid_cell_area_ESMs"

#Getting a list of sample rasters in the isimip folders
sample_rasters <- list.files(samples_dir, pattern = ".nc$", full.names = T,
                             recursive = T) |> 
  str_subset("isimip")

#Checking results
sample_rasters
```
  
We will define a function that will go through each sample file and create a mask.  
  
```{r}
shp_to_raster <- function(shp, raster_path, field, name_sufix, out_folder){
  #Transforming shapefile to SpatVector class
  shp_terra <- vect(shp)
  #Loading raster
  ras <- rast(raster_path)
  #Rasterise shapefile
  shp_rast <- rasterize(shp_terra, ras, field = field, background = NA)
  #Create name for mask to be saved from original raster sample
  file_out <- str_replace(basename(raster_path), "global_fixed", name_sufix)
  file_out <- file.path(out_folder, file_out)
  writeCDF(shp_rast, file_out, overwrite = T, varname = "region", missval = NA)
}
```
  
Applying function to all raster samples.  
  
```{r, eval = F}
#Ensure output folder exists
out_folder <- "/rd/gem/private/shared_resources/SouthernOceanMasks"
if(!dir.exists(out_folder)){
  dir.create(out_folder, recursive = T)}

#Applying function creating masks
for(ras in sample_rasters){
  shp_to_raster(southern_ocean, ras, "id", "Southern_Ocean_mask", out_folder)
}
```
  
# Plotting mask
We will plot one mask to ensure it has been correctly created.  
  
```{r}
ras <- rast(list.files("/rd/gem/private/shared_resources/SouthernOceanMasks/", 
                       "w-fractions", full.names = T))
plot(ras)
```
  
The final mask matches the region, which is exactly what we needed. We are now going to save a data frame containing the name of the Southern Ocean regions and their unique ID so they can be easily interpreted.  
  
```{r, eval = FALSE}
southern_ocean |> 
  st_drop_geometry() |> 
  distinct() |> 
  write_csv(file.path(out_folder, "Southern_Ocean_region_keys.csv"))
```
  
We can now use these masks to extract data from any rasters sharing the same grid as the mask.  
  
      
## Creating a simple 2D mask for all FishMIP regions
  
```{r}
fishmip_reg <- read_sf(file.path("/rd/gem/private/shared_resources", 
                                 "FishMIP_regional_models", 
                                 "FishMIP_regional_models.shp"))

out_folder <- file.path("/rd/gem/private/shared_resources/FishMIPMasks",
                        "merged_regional_fishmip")
if(!dir.exists(out_folder)){
  dir.create(out_folder, recursive = T)}

#Applying function creating masks
for(ras in sample_rasters){
  shp_to_raster(fishmip_reg, ras, 1, "fishMIP_regional_merged", out_folder)
}
```

