---
title: "Creating_multidimensional_raster_mask"
author: "Denisse Fierro Arcos"
date: "2023-11-27"
output:
  github_document:
    toc: true
    html_preview: false
---

## Introduction

In this notebook, we will show how to create a multidimensional raster mask that can be used to extract data from Earth System Models from the Fish-MIP regional models shapefile. We create a multidimensional mask because there are multiple overlapping regions. We will create three-dimensional masks that will match the various grids used in the Fish-MIP project, as well as a mask in table format (i.e., data frame).  
  
# Loading libraries
  
```{r libraries, warning = F, message = F}
#Spatial data
library(sf)
library(terra)
library(stars)

#Manipulating and plotting data
library(tidyverse)

#Base map
library(rnaturalearth)
```
  
# Loading Fish-MIP regional models shapefile
We will load the shapefile with all Fish-MIP regional model boundaries and add a unique ID identifying each region.  
  
```{r}
#Loading shapefile
regions <- file.path("/rd/gem/private/shared_resources/",
                     "FishMIP_regional_models/FishMIP_regional_models.shp") |> 
  read_sf() |> 
  #Create a unique ID for each region
  rowid_to_column("id")

#Checking result
ne_countries(returnclass = "sf") |> 
  ggplot()+
  geom_sf()+
  geom_sf(inherit.aes = F, data = regions, aes(fill = region), alpha = 0.7)+
  theme_bw()
```
  
Our shapefile is plotting correctly, now we can move onto creating raster masks. In total, we will create four different multidimensional raster masks to match all data available as forcings for Fish-MIP models.  
  
# Finding location of raster samples
We have a folder containing samples of the raster used in Fish-MIP models and ESMs. We will list all the files contained in that folder.    
  
```{r}
#Directory containing sample rasters
samples_dir <- "/rd/gem/private/shared_resources/grid_cell_area_ESMs"

#Getting a list of sample rasters in the isimip folders
sample_rasters <- list.files(samples_dir, pattern = ".nc$", full.names = T,
                             recursive = T) |> 
  str_subset("isimip")

#Checking results
sample_rasters
```
  
We will define a function that will go through each sample file and create a mask.  
  
```{r}
shp_to_raster <- function(shp, raster_path){
  #Transforming shapefile to SpatVector class
  shp_terra <- vect(shp)
  #Loading raster
  ras <- rast(raster_path)
  #Rasterise shapefile
  shp_rast <- rasterize(shp_terra, ras, field = "id", background = NA)
  return(shp_rast)
}
```
  
Before applying the function, we will make sure that an output folder exists, if not, we will create one.  
  
```{r}
#Ensure output folder exists
out_folder <- "/rd/gem/private/shared_resources/FishMIPMasks"
if(!dir.exists(out_folder)){
  dir.create(out_folder, recursive = T)}

#Storing each region inside the regions shapefile as an element of a list
region_list <- split(regions, 1:nrow(regions))
```
  
Applying function to all raster samples.  
  
```{r, eval = F}
#Applying function creating masks
for(ras in sample_rasters){
  stack_list <- map(region_list, shp_to_raster, ras)
  stack <- rast(stack_list)
  #Create name for mask to be saved from original raster sample
  file_out <- str_replace(basename(ras), 
                          "global", "fishMIP_regional_mask")
  file_out <- file.path(out_folder, file_out)
  #Save multi dimensional raster mask
  writeCDF(stack, file_out, overwrite = T, varname = "region",
           longname = "Fish-MIP regions", missval = NA)
}
```
  
# Plotting mask
We will plot one mask to ensure it has been correctly created.  
  
```{r}
ras <- rast(list.files("/rd/gem/private/shared_resources/FishMIPMasks", 
                       "w-fractions.*nc", full.names = T))
plot(ras)
```
  
# Saving raster mask keys
We will now save a data frame containing the name of the Fish-MIP regions and their unique ID so the regions inside the masks can be easily interpreted.  
  
```{r}
reg_keys <- regions |> 
  #Remove geometry
  st_drop_geometry() |> 
  #Keep region names and ID
  select(id, region) 

#Save keys in masks folder
reg_keys |> 
  write_csv(file.path(out_folder, "FishMIP_regions_keys.csv"))

#Check keys
reg_keys
```
  
## How to use raster mask
If the ESM data is in raster form, we can transform the raster mask into a binary mask to extract the data we need.  
  
```{r}
#Load sample ESM data
sample <- rast(str_subset(sample_rasters, "w-fractions"))

#We will choose mask 9 - East Antarctica (see keys above)
east_ant <- ras[[9]]
#We will replace the ID for the region for the value of 1
east_ant[!is.na(east_ant)] = 1

#Multiply data and mask
extract_data <- sample*east_ant
#Check result
plot(extract_data)
```
  
# Data frame mask
We will also create a data frame matching all the raster masks that were created above. First, we define a function that will transform `netCDF` masks to `csv` masks.  
  
```{r}
mask_df <- function(raster_path, keys){
  df <- rast(raster_path) |> 
    as.data.frame(xy = T) |> 
    rename("lon" = "x", "lat" = "y") |> 
    pivot_longer(starts_with("region"), names_to = "region", values_to = "id", 
                 values_drop_na = T) |>
    select(!region) |> 
    left_join(keys, by = "id")
  
  #Create name for mask to be saved from original raster sample
  file_out <- str_replace(raster_path, ".nc", ".csv")
  #Save file
  write_csv(df, file_out)
}
```
  
Now we will apply this function to all raster masks.  
  
```{r, eval = F}
raster_masks <- list.files(out_folder, pattern = ".nc$", full.names = T)

#Applying function creating csv masks
for(ras in raster_masks){
  mask_df(ras, reg_keys)}
```
  
## How to use data frame mask
If the ESM data is in data frame form, we can use the `csv` mask to extract the data we need.  
  
```{r}
#Load sample ESM data
sample_df <- sample |> 
  #Transforming to data frame
  as.data.frame(xy = T) |> 
  rename("lon" = "x", "lat" = "y")

#Load raster mask
mask_df <- read_csv(list.files(out_folder, "w-fractions.*csv",
                               full.names = T)) |> 
  #We will choose mask 9 - East Antarctica (see keys above)
  filter(region == "East Antarctica EwE")

#Apply mask to ESM data
extract_df <- mask_df |> 
  left_join(sample_df, by = c("lon", "lat"))

#Plotting result
extract_df |> 
  ggplot()+
  geom_raster(aes(x = lon, y = lat, fill = areacello))
```
  
You can also apply the mask to extract all data at once.  
  
```{r}
#Apply mask to ESM data
extract_df_all <- read_csv(list.files(out_folder, "w-fractions.*csv",
                               full.names = T)) |> 
  left_join(sample_df, by = c("lon", "lat"))

#Plotting result
extract_df_all |> 
  ggplot()+
  #Color by region, transparency by area
  geom_raster(aes(x = lon, y = lat, fill = region, alpha = areacello))
```
  
**Note:** When you use a mask, whether in raster or `csv` form, the grid for the mask and the data being extracted **MUST** be the same. Otherwise, data extraction will fail.  
  
